Supporting Information for\\ ``A heterogeneous landscape of selection and epistasis in protein-coding genes revealed by two-locus statistics''

### 2022-04-01 manuscript/paper.tex
ready to resubmit

@@

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/two-locus-spectrum/stats_summaries}
    \caption{
        \textbf{Sampling distributions and their summaries.}
        Low-order summaries of sampling distributions are commonly computed
        for allele frequencies (A, the site-frequency spectrum, SFS)
        and two-locus haplotype distributions (B, linkage disequilibrium, LD).
        Demographic and selective processes affect both the SFS and LD,
        and observations of nonzero values of Tajima's $D$ or signed LD
        ($\sigma_d^1$) are often taken as evidence for selection or interactions
        between loci, respectively.
        (C) The full two-locus haplotype sampling distribution is a three-dimensional
        object, making it difficult to visualize. We can instead visualize conditional
        distributions of the full sampling distribution, e.g., conditioned on observing
        \(n_A\) copies of the \(A\) allele at the left locus and \(n_B\) copies of
        \(B\) at the right locus \citep[e.g.,][]{Hudson2001-sg}.
        (D) $\sigma_d^2$, which is closely related to $r^2$, is expected to decay
        with increasing recombination distance between loci \citep{Ohta1969-ie}.
        Selection distorts squared LD away from neutral expectations.
        (E) $\sigma_d^1$, refered to as signed LD here, is expected to be zero for
        pairs of neutral mutations \citep{Hill1968-vu}. Interference between linked
        selected mutation causes negative signed LD \citep{Hill1966-gv}, and other
        forms of interactions between selected mutations can cause large negative
        or positive signed LD.
    }
    \label{fig:TwoLocusFS}
\end{figure}

@@

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/hill_robertson}
    \caption{
        \textbf{Hill-Robertson interference.}
        Interference between pairs of selected mutations can cause negative
        signed LD in the absence of dominance and epistatic effects.
        (A) The expected normalized variance of $D$ (\(\sigma_d^2\))
        decreases below neutral expectations
        as the strength of negative selection increases.
        (B, C) For tightly linked loci (\(4Nr \lesssim 1\)), interference is most
        noticeable for pairs of mutations with \(s \approx 1/N\).
        At larger recombination distances (\(4Nr > 1\)), signed LD is most
        negative for somewhat stronger selection coefficients.
        Dashed lines show neutral expectations.
    }
    \label{fig:HillRobertson}
\end{figure}

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/epistasis_prediction}
    \caption{
        \textbf{Additive selection and epistasis.}
        Left panels (A and C) show expectations for the decay of
        \(\sigma_d^2 = \E[D^2] / \E[p(1-p)q(1-q)]\) with recombination
        distance, and right panels (B and D) show expectations for the
        decay of \(\sigma_d^1 = \E[D] / \E[p(1-p)q(1-q)]\). Dashed lines
        show neutral expectations.
        For both weak (\(s=-1/2N\), A and B) and moderate
        (\(s=-10/2N\), C and D) selection,
        antagonistic epistasis (\(\epsilon < 0\))
        gives rise to positive signed LD and increased $\sigma_d^2$ over
        a multiplicative model (\(\epsilon = 0\)),
        and synergistic epistasis (\(\epsilon > 0\)) results in negative signed
        LD beyond that of Hill-Robertson interference and decreased \(\sigma_d^2\).
    }
    \label{fig:epistasis}
\end{figure}

@@

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/dominance_roze}
    \caption{
        \textbf{The effect of dominance on LD.}
        (A, B) The strengths of selection and dominance interact in a nonlinear way
        to shape expected signed LD. For weakly to moderately selected mutations,
        as shown here, signed LD can be large and negative for tightly linked loci
        (e.g., \(\gamma=-5\), \(h=0\), and \(\rho < 1\)). However, this large signed
        LD decays with recombination distance faster under a model of recessivity
        than does signed LD under a model of additive selection and
        epistasis (Figure~\ref{fig:epistasis}).
        (C) Interference effects are most pronounced for recessive deleterious
        variants.
        (D, E) Recessive strongly deleterious mutations can have positive signed
        LD, as recently shown by \citet{Roze2021-cf}.
        However the dominance threshold at which \(\sigma_d^1\) switches from
        positive to negative depends on the strength of selection, and weakly
        selected mutations can show non-monotonic behavior as $h$ varies.
        Selection parameters of \(sh=-0.1\) imply extremely strong selection
        (\(h=0.05\) results in \(s=-0.2\) and \(\gamma=-400\) for homozygous
        diploids at a single locus). The numerical appraoch for \(\Psi_n\)
        cannot handle such strong selection.
    }
    \label{fig:dominance}
\end{figure}

@@

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/gene-based-compensatory}
    \caption{
        \textbf{Multiple modes of interactions can lead to large positive signed LD.}
        Both antagonistic epistasis (A and C, and which includes compensatory mutation
        models) and gene-based dominance (B and D) lead to large positive signed LD.
        Compensatory mutations (\(\epsilon \approx -1\)) also cause increased
        \(\sigma_d^2\) compared to neutral expectations (dashed black lines),
        while weaker antagonistic epistasis does not increase \(\sigma_d^2\) above
        neutral expectations (compare to Figure~\ref{fig:epistasis}C).
        Gene-based dominance instead causes lower \(\sigma_d^2\) than neutral
        expectations.
        While signed LD may be similar between different interaction models,
        other two-locus summaries of the data may help to distinguish between
        such models.
        }
    \label{fig:gene-based-compensatory}
\end{figure}

@@

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/demog_bottle_expand}
    \caption{
        \textbf{The effects of demography on signed LD.}
        Simulations under two toy models, one with an instantaneous expansion in
        the past and the other with a five-fold reduction and recovery illustrated
        in (A), show the effects of population size changes on expected patterns of LD. 
        For two selection strengths (\(\gamma=-1\) and \(\gamma=-10\)) and three
        cases of interactions (non-epistatic interference (\(\epsilon=0\), B),
        synergistic epistasis (\(\epsilon=1/2\), C), and antagonistic epistasis
        (\(\epsilon=-1/2\), D), each with \(\rho=0\)),
        sudden decreases in population size can cause
        large changes in signed LD, often in the opposite direction than more
        subtle shifts due to instantaneous expansion events.
        (E) Signed LD between pairs of rare alleles (both \(n_A, n_B \leq 4\), with
        sample size \(n=50\)) is more sensitive to population size changes.
        (F) Signed LD between pairs of common alleles (both \(n_A, n_B \geq 5/50\))
        is comparatively more stable over time.
        Additional comparisons, including for dominance models and showing
        \(\sigma_d^2\), are shown in Figures~\ref{fig:toy_sd2}--\ref{fig:toy_dom_sd2}.
        Dashed lines indicate neutral expectations.
    }
    \label{fig:toy}
\end{figure}

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/demog_YRI_CEU.dominance}
    \caption{
        \textbf{Signed LD under inferred models of human population-size history.}
        Piecewise-constant population size histories inferred by \texttt{Relate} applied to
        \citet{1000_Genomes_Project_Consortium2015-zq} phase 3 data were used to
        simulate time series of two-locus statistics, as in Figure~\ref{fig:toy}.
        (A) The CEU (Utah residents with Northern and Western European ancestry)
        are inferred to have a stronger bottleneck than the YRI
        (Yoruba in Ibadan, Nigeria) 10--100ka, reflecting the out-of-Africa
        event.
        Here, we highlight the effect of size changes on \(\sigma_d^1\) under
        non-additive selection models with \(\rho=0\) and \(\gamma=-2\) at both loci,
        comparing standard interference (B) to site-wise dominance (C) and gene-based
        dominance (C).
        As with epistasis, more severe bottlenecks have larger effects on signed LD,
        and LD among common variants is more stable than among pairs of rare variants.
        Additional comparisons with epistasis and showing \(\sigma_d^2\) are
        in Figures~\ref{fig:relate}--\ref{fig:relate_dom_sd2}.
        Dashed lines indicate neutral expectations.
    }
    \label{fig:relate_dom}
\end{figure}

@@

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/data_compact}
    \caption{
        \textbf{LD in human protein-coding genes and annotated domains.}
        (A) Gene-wide averages of signed LD are slightly positive between pairs
        of both missense and synonymous mutations, considering pairs of mutations
        at matching distances. This positive, equal \(\sigma_d^1\) is also observed
        when conditioning on allele frequencies or considering only common variants
        with minor allele frequencies \(\gtrsim 0.1\)
        (Tables~\ref{tab:afrsd1}--\ref{tab:eassd2}).
        (B) While there are relatively fewer pairs of loss-of-function mutations
        within genes, causing larger measurement uncertainty, such pairs tend to have
        negative average LD. Note that measurement noises for each class of
        mutations overlap with zero and with each other, making it difficult to
        draw firm conclusions on the patterns of interactions occurring gene-wide.
        (C) Partitioning pairs of mutations as falling within or outside
        of conserved domains reveals opposing patterns of signed LD, with
        \(\sigma_d^1\) between missense mutations larger than that of synonymous
        mutations within domains. Outside of conserved domains, missense mutations
        have reduced LD compared to synonymous mutations. Distances of pairs outside
        of domains were matched to within-domain mutation pair distances.
        (D) This pattern again holds for common variants.
        (E, F) The signal of increased LD between missense variants within domains
        and decreased LD outside of domains is driven largely by tightly linked
        mutations (distances \(\lesssim 400\) bp), here showing African populations
        in the \citet{1000_Genomes_Project_Consortium2015-zq}.
        Additional comparisons are shown in the Supporing Information, and Thousand
        Genomes population labels are described in Table~\ref{tab:1kgpops}.
    }
    \label{fig:data}
\end{figure}

@@

I thank Alex Diaz-Papkovich, Eric Friedlander, Simon Gravel, Carol Lee, Mashaal
Sohail, Matthias Steinrücken, and Kevin Thornton for helpful discussions and
feedback on earlier versions of this manuscript.

>>>>

\usepackage{lineno}
\linenumbers

@@

\begin{figure}[tb!]
    \internallinenumbers
    \centering
    \includegraphics{../figures/two-locus-spectrum/stats_summaries}
    \caption{
        \textbf{Sampling distributions and their summaries.}
        Low-order summaries of sampling distributions are commonly computed
        for allele frequencies (A, the site-frequency spectrum, SFS)
        and two-locus haplotype distributions (B, linkage disequilibrium, LD).
        Demographic and selective processes affect both the SFS and LD,
        and observations of nonzero values of Tajima's $D$ or signed LD
        ($\sigma_d^1$) are often taken as evidence for selection or interactions
        between loci, respectively.
        (C) The full two-locus haplotype sampling distribution is a three-dimensional
        object, making it difficult to visualize. We can instead visualize conditional
        distributions of the full sampling distribution, e.g., conditioned on observing
        \(n_A\) copies of the \(A\) allele at the left locus and \(n_B\) copies of
        \(B\) at the right locus \citep[e.g.,][]{Hudson2001-sg}.
        (D) $\sigma_d^2$, which is closely related to $r^2$, is expected to decay
        with increasing recombination distance between loci \citep{Ohta1969-ie}.
        Selection distorts squared LD away from neutral expectations.
        (E) $\sigma_d^1$, refered to as signed LD here, is expected to be zero for
        pairs of neutral mutations \citep{Hill1968-vu}. Interference between linked
        selected mutation causes negative signed LD \citep{Hill1966-gv}, and other
        forms of interactions between selected mutations can cause large negative
        or positive signed LD.
    }
    \label{fig:TwoLocusFS}
\end{figure}

@@

\begin{figure}[tb!]
    \internallinenumbers
    \centering
    \includegraphics{../figures/hill_robertson}
    \caption{
        \textbf{Hill-Robertson interference.}
        Interference between pairs of selected mutations can cause negative
        signed LD in the absence of dominance and epistatic effects.
        (A) The expected normalized variance of $D$ (\(\sigma_d^2\))
        decreases below neutral expectations
        as the strength of negative selection increases.
        (B, C) For tightly linked loci (\(4Nr \lesssim 1\)), interference is most
        noticeable for pairs of mutations with \(s \approx 1/N\).
        At larger recombination distances (\(4Nr > 1\)), signed LD is most
        negative for somewhat stronger selection coefficients.
        Dashed lines show neutral expectations.
    }
    \label{fig:HillRobertson}
\end{figure}

\begin{figure}[tb!]
    \internallinenumbers
    \centering
    \includegraphics{../figures/epistasis_prediction}
    \caption{
        \textbf{Additive selection and epistasis.}
        Left panels (A and C) show expectations for the decay of
        \(\sigma_d^2 = \E[D^2] / \E[p(1-p)q(1-q)]\) with recombination
        distance, and right panels (B and D) show expectations for the
        decay of \(\sigma_d^1 = \E[D] / \E[p(1-p)q(1-q)]\). Dashed lines
        show neutral expectations.
        For both weak (\(s=-1/2N\), A and B) and moderate
        (\(s=-10/2N\), C and D) selection,
        antagonistic epistasis (\(\epsilon < 0\))
        gives rise to positive signed LD and increased $\sigma_d^2$ over
        a multiplicative model (\(\epsilon = 0\)),
        and synergistic epistasis (\(\epsilon > 0\)) results in negative signed
        LD beyond that of Hill-Robertson interference and decreased \(\sigma_d^2\).
    }
    \label{fig:epistasis}
\end{figure}

@@

\begin{figure}[tb!]
    \internallinenumbers
    \centering
    \includegraphics{../figures/dominance_roze}
    \caption{
        \textbf{The effect of dominance on LD.}
        (A, B) The strengths of selection and dominance interact in a nonlinear way
        to shape expected signed LD. For weakly to moderately selected mutations,
        as shown here, signed LD can be large and negative for tightly linked loci
        (e.g., \(\gamma=-5\), \(h=0\), and \(\rho < 1\)). However, this large signed
        LD decays with recombination distance faster under a model of recessivity
        than does signed LD under a model of additive selection and
        epistasis (Figure~\ref{fig:epistasis}).
        (C) Interference effects are most pronounced for recessive deleterious
        variants.
        (D, E) Recessive strongly deleterious mutations can have positive signed
        LD, as recently shown by \citet{Roze2021-cf}.
        However the dominance threshold at which \(\sigma_d^1\) switches from
        positive to negative depends on the strength of selection, and weakly
        selected mutations can show non-monotonic behavior as $h$ varies.
        Selection parameters of \(sh=-0.1\) imply extremely strong selection
        (\(h=0.05\) results in \(s=-0.2\) and \(\gamma=-400\) for homozygous
        diploids at a single locus). The numerical appraoch for \(\Psi_n\)
        cannot handle such strong selection.
    }
    \label{fig:dominance}
\end{figure}

@@

\begin{figure}[tb!]
    \internallinenumbers
    \centering
    \includegraphics{../figures/gene-based-compensatory}
    \caption{
        \textbf{Multiple modes of interactions can lead to large positive signed LD.}
        Both antagonistic epistasis (A and C, and which includes compensatory mutation
        models) and gene-based dominance (B and D) lead to large positive signed LD.
        Compensatory mutations (\(\epsilon \approx -1\)) also cause increased
        \(\sigma_d^2\) compared to neutral expectations (dashed black lines),
        while weaker antagonistic epistasis does not increase \(\sigma_d^2\) above
        neutral expectations (compare to Figure~\ref{fig:epistasis}C).
        Gene-based dominance instead causes lower \(\sigma_d^2\) than neutral
        expectations.
        While signed LD may be similar between different interaction models,
        other two-locus summaries of the data may help to distinguish between
        such models.
        }
    \label{fig:gene-based-compensatory}
\end{figure}

@@

\begin{figure}[tb!]
    \internallinenumbers
    \centering
    \includegraphics{../figures/demog_bottle_expand}
    \caption{
        \textbf{The effects of demography on signed LD.}
        Simulations under two toy models, one with an instantaneous expansion in
        the past and the other with a five-fold reduction and recovery illustrated
        in (A), show the effects of population size changes on expected patterns of LD. 
        For two selection strengths (\(\gamma=-1\) and \(\gamma=-10\)) and three
        cases of interactions (non-epistatic interference (\(\epsilon=0\), B),
        synergistic epistasis (\(\epsilon=1/2\), C), and antagonistic epistasis
        (\(\epsilon=-1/2\), D), each with \(\rho=0\)),
        sudden decreases in population size can cause
        large changes in signed LD, often in the opposite direction than more
        subtle shifts due to instantaneous expansion events.
        (E) Signed LD between pairs of rare alleles (both \(n_A, n_B \leq 4\), with
        sample size \(n=50\)) is more sensitive to population size changes.
        (F) Signed LD between pairs of common alleles (both \(n_A, n_B \geq 5/50\))
        is comparatively more stable over time.
        Additional comparisons, including for dominance models and showing
        \(\sigma_d^2\), are shown in Figures~\ref{fig:toy_sd2}--\ref{fig:toy_dom_sd2}.
        Dashed lines indicate neutral expectations.
    }
    \label{fig:toy}
\end{figure}

\begin{figure}[tb!]
    \internallinenumbers
    \centering
    \includegraphics{../figures/demog_YRI_CEU.dominance}
    \caption{
        \textbf{Signed LD under inferred models of human population-size history.}
        Piecewise-constant population size histories inferred by \texttt{Relate} applied to
        \citet{1000_Genomes_Project_Consortium2015-zq} phase 3 data were used to
        simulate time series of two-locus statistics, as in Figure~\ref{fig:toy}.
        (A) The CEU (Utah residents with Northern and Western European ancestry)
        are inferred to have a stronger bottleneck than the YRI
        (Yoruba in Ibadan, Nigeria) 10--100ka, reflecting the out-of-Africa
        event.
        Here, we highlight the effect of size changes on \(\sigma_d^1\) under
        non-additive selection models with \(\rho=0\) and \(\gamma=-2\) at both loci,
        comparing standard interference (B) to site-wise dominance (C) and gene-based
        dominance (C).
        As with epistasis, more severe bottlenecks have larger effects on signed LD,
        and LD among common variants is more stable than among pairs of rare variants.
        Additional comparisons with epistasis and showing \(\sigma_d^2\) are
        in Figures~\ref{fig:relate}--\ref{fig:relate_dom_sd2}.
        Dashed lines indicate neutral expectations.
    }
    \label{fig:relate_dom}
\end{figure}

@@

\begin{figure}[tb!]
    \internallinenumbers
    \centering
    \includegraphics{../figures/data_compact}
    \caption{
        \textbf{LD in human protein-coding genes and annotated domains.}
        (A) Gene-wide averages of signed LD are slightly positive between pairs
        of both missense and synonymous mutations, considering pairs of mutations
        at matching distances. This positive, equal \(\sigma_d^1\) is also observed
        when conditioning on allele frequencies or considering only common variants
        with minor allele frequencies \(\gtrsim 0.1\)
        (Tables~\ref{tab:afrsd1}--\ref{tab:eassd2}).
        (B) While there are relatively fewer pairs of loss-of-function mutations
        within genes, causing larger measurement uncertainty, such pairs tend to have
        negative average LD. Note that measurement noises for each class of
        mutations overlap with zero and with each other, making it difficult to
        draw firm conclusions on the patterns of interactions occurring gene-wide.
        (C) Partitioning pairs of mutations as falling within or outside
        of conserved domains reveals opposing patterns of signed LD, with
        \(\sigma_d^1\) between missense mutations larger than that of synonymous
        mutations within domains. Outside of conserved domains, missense mutations
        have reduced LD compared to synonymous mutations. Distances of pairs outside
        of domains were matched to within-domain mutation pair distances.
        (D) This pattern again holds for common variants.
        (E, F) The signal of increased LD between missense variants within domains
        and decreased LD outside of domains is driven largely by tightly linked
        mutations (distances \(\lesssim 400\) bp), here showing African populations
        in the \citet{1000_Genomes_Project_Consortium2015-zq}.
        Additional comparisons are shown in the Supporing Information, and Thousand
        Genomes population labels are described in Table~\ref{tab:1kgpops}.
    }
    \label{fig:data}
\end{figure}

@@

I thank Alex Diaz-Papkovich, Eric Friedlander, Simon Gravel, Mashaal Sohail,
Matthias Steinrücken, and Kevin Thornton for helpful discussions and feedback
on earlier versions of this manuscript.

### 2022-04-01 manuscript/supplement.tex
ready to resubmit

\title{Supporting Information for\\
``A heterogeneous landscape of selection and epistasis in protein-coding genes''}
\author[]{Aaron P. Ragsdale}
\affil[1]{Department of Integrative Biology, University of Wisconsin, Madison, WI, USA}
\affil[*]{apragsdale@wisc.edu}
\date{\today}
\maketitle

@@

Simulations had diploid population sizes of 5000 and I drew haploid sample
sizes of 50 (25 diploids), with recombination rates varying from \(0\) to
\(0.001\) (or \(\rho=0-20\)). I tested selection coefficients of \(s\) of 0,
\(-1\times10^{-5}\), \(-2\times10^{-4}\), and \(-2\times10^{-3}\) (or
\(\gamma=0\), \(-0.1\), \(-2\), and \(-20\)). Epistasis was set to
\(\epsilon=0\), \(-0.5\), and \(0.5\), and dominance was set to either
\(h=0.5\) (additive) or \(0.1\) (recessive). Conditional distributions of the
two-locus sampling distribution show good agreement between numerical solutions
and simulations, and residuals do not show any directional bias
(Figures~\ref{fig:validation}--\ref{fig:last_validation}). The Support Figures
includes a handful of these comparisons, and the full set of comparisons (over
40 in total across the range of parameters listed above), simulation, and
plotting scripts in Python are available at
\url{https://github.com/apragsdale/two_locus_selection}.

>>>>

\title{Supporting Information for\\
``A heterogeneous landscape of selection and epistasis in protein-coding genes revealed by two-locus statistics''}
\author[]{Aaron P. Ragsdale}
\affil[]{Department of Integrative Biology, University of Wisconsin, Madison, WI, USA}
\affil[]{apragsdale@wisc.edu}
\date{\today}
\maketitle

@@

Simulations had diploid population sizes of 5000 and I drew haploid sample
sizes of 50 (25 diploids), with recombination rates varying from \(0\) to
\(0.001\) (or \(\rho=0-20\)). I tested selection coefficients of \(s\) of 0,
\(-1\times10^{-5}\), \(-2\times10^{-4}\), and \(-2\times10^{-3}\) (or
\(\gamma=0\), \(-0.1\), \(-2\), and \(-20\)). Epistasis was set to
\(\epsilon=0\), \(-0.5\), and \(0.5\), and dominance was set to either
\(h=0.5\) (additive) or \(0.1\) (recessive). Conditional distributions of the
two-locus sampling distribution show good agreement between numerical solutions
and simulations, and residuals do not show any directional bias
(Figures~\ref{fig:validation1}--\ref{fig:validation5}). The Supporting Figures
includes a handful of these comparisons, and the full set of comparisons (over
40 in total across the range of parameters listed above), simulation, and
plotting scripts in Python are available at
\url{https://github.com/apragsdale/two_locus_selection}.
